Plug-and-play dual-tree algorithm runtime analysis

### 2015-10-30 bounds-files/general-runtime-bound.tex
Particularly trivial revisions.

Now suppose that $s_i < s_r^{\max} - 1$.  This implies that there exists some
implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well an
implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth until
one of these implicit nodes has child $p_i$ with scale $s_i$).  Because the
separation invariant applies to both implicit and explicit representations of
the tree, we conclude that $d(p_i, p_j) > 2^{s_r^{\max}} - 1$.  The same
argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same
conclusion.

We may therefore conclude that each point of each node in $R_{r - 1}$ is
separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
and that $R \setminus R_{r - 1} \subseteq R$ in order to see that this condition
holds for all nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.

Because we have shown that the condition holds for the initial reference set and
for any reference set produced by a reference recursion, we have shown that the
statement of the lemma is true.

@@

\begin{defn}
Define $\theta$ as a bound on the number of extra reference recursions tha
happen after the last query recursion.  Then,

@@

\begin{proof}
First, split the algorithm into two parts: reference recursions (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
and query recursions (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
The runtime of the algorithm is bounded as the runtime of a reference recursion
times the total number of reference recursions plus the total runtime of all
query recursions.

@@

When we consider the monochromatic case (where $S_q = S_r$)\footnote{In the
monochromatic case, we do not take a point as its own nearest neighbor, so
slight modification of {\tt BaseCase()} is necessary.  The runtime bound result
remains unchanged.}, the results trivially simplify.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
then the query tree must be entirely descended before any reference recursion.
This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
$O(N^2)$ (See Lemma \ref{lem:extcase3}).  This is because the reference tree
must be descended individually for each query point.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (Lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work.  An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics \citep{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
can be bounded no more tightly than $O(N)$ with our bound, which matches the
intuition that answering $O(N)$ queries will take at least $O(N)$ time. For a
particular problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by
constants independent of $N$ and $\theta$ is no more than linear in $N$ (for
large enough $N$), then the dual-tree algorithm for that problem has a runtime
linear in $N$.  Our theoretical result separates out the problem-dependent and
the problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds in order to get runtime bounds for any
dual-tree algorithm without requiring an analysis from scratch.

>>>>

Now suppose that $s_i < s_r^{\max} - 1$.  This implies that there exists some
implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well as
an implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth
until one of these implicit nodes has child $p_i$ with scale $s_i$).  Because
the separation invariant applies to both implicit and explicit representations
of the tree, we conclude that $d(p_i, p_j) > 2^{s_r^{\max}} - 1$.  The same
argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same
conclusion.

We may therefore conclude that each point of each node in $R_{r - 1}$ is
separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
and that $R \setminus R_{r - 1} \subseteq R$ in order to see that this condition
holds for all nodes in $R'_{r - 1}$.

Because we have shown that the condition holds for the initial reference set and
for any reference set produced by a reference recursion (which will be $R$ at
some other level of recursion), we have shown that the statement of the lemma is
true.

@@

\begin{defn}
Define $\theta$ as a bound on the number of extra reference recursions that
happen after the last query recursion.  Then,

@@

\begin{proof}
First, split the algorithm into two parts: reference recursions (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
and query recursions (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
The runtime of the algorithm is the runtime of a reference recursion
times the total number of reference recursions plus the total runtime of all
query recursions.

@@

When we consider the monochromatic case (where $S_q = S_r$), the results trivially simplify.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
then the query tree must be entirely descended before any reference recursion.
This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
$O(N^2)$ (see Lemma \ref{lem:extcase3}).  This is because the reference tree
must be descended separately for each query point.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work.  An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics \citep{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
can be bounded no more tightly than $O(N)$ with our bound, which matches the
intuition that answering $O(N)$ queries will take at least $O(N)$ time. For a
particular problem and data, if $c_r$, $|R^*|$, $\chi$, and $\psi$ are bounded by
constants independent of $N$ and $\theta$ is no more than linear in $N$ (for
large enough $N$), then the dual-tree algorithm for that problem has a runtime
linear in $N$.  Our theoretical result separates out the problem-dependent and
the problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds in order to get runtime bounds for any
dual-tree algorithm without requiring an analysis from scratch.

### 2015-10-30 bounds-files/imbalance.tex
Add higgs dataset.
Particularly trivial revisions.

\begin{figure}
\begin{subfigure}[b]{0.585\textwidth}
  \begin{center}
    \input{figures/balanced-cover-tree.tex}
  \end{center}
  \caption{Balanced cover tree.}
  \label{fig:imbalance-good}
\end{subfigure}
\begin{subfigure}[b]{0.415\textwidth}
  \begin{center}
    \input{figures/imbalanced-cover-tree.tex}
  \end{center}
  \caption{Imbalanced cover tree.}
  \label{fig:imbalance-bad}
\end{subfigure}
\caption{Balanced and imbalanced cover trees.}
\label{fig:imbalance}
\end{figure}

@@

\begin{defn}
The {\it cover node imbalance} $I_n(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf child (that is, $s_i = -\infty$), then number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $I_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $I_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
this situation, $c \sim N$ also.}.

However, for most real-world datasets with the cover tree implementation in {\bf
mlpack} \citep{mlpack2013} and the reference implementation
\citep{langford2006}, the tree imbalance is near-linear with the number of
points.  We have constructed cover trees on $N$ uniformly subsampled points
from a variety of datasets and calculated the imbalance; see Table
\ref{tab:imbalance} for the results.  Ten trials were performed for each dataset
and each $N$, and the mean imbalance is given.  These results are normalized
with respect to $N$, for which the values of $5000$, $50000$, and $500000$ were
chosen.  The `power', `susy', `higgs', and `covertype' datasets are found in the
UCI Machine Learning Repository \cite{ucimlrepository}, the `mnist' dataset is
from \cite{lecun2000mnist}, the `lcdm` and `sdss' datasets are Sloan Digital Sky
Survey data \cite{adelman2008sixth}, and the `randu' dataset is
randomly-generated uniformly-distributed data in 10 dimensions.  The imbalances
on each of these datasets tend to be near-linear.

@@

\begin{table}[tb]
\begin{center}
\begin{tabular}{| c | c | c | c | c |}
\hline
 & & \multicolumn{3}{|c|}{Imbalance} \\
\hline
Dataset & $d$ & $N = 5k$ & $N = 50k$ & $N = 500k$ \\
\hline
%lcdm & 3 & 22417.9 & 257445.6 & 2621239.6 \\
lcdm & 3 & 4.48 & 5.15 & 5.24 \\
% sdss & 4 & 10853.6 & 140431.6 & 1486808.8 \\
sdss & 4 & 2.17 & 2.81 & 2.97 \\
% power & 7 & 27042.2 & 323125.6 & 2251317.2
power & 7 & 5.41 & 6.46 & 4.50 \\
% susy & 18 & 3701.3 & 37823.7 & 431784.8 \\
susy & 18 & 0.74 & 0.76 & 0.86 \\
%randu-10 & 10 & 1156.0 & 10763.4 & 297041.2 \\
randu & 10 & 0.23 & 0.22 & 0.59 \\
% higgs & 29 & 4963.6 & 84171.8 &  \\ need one more...
%covertype & 54 & 6609.6 & 88296.4 & 1247820.7 \\
covertype & 54 & 1.322 & 1.766 & 2.495 \\
% mnist & 784 & 4974.1 & 83658.6 & 1044075.0 \\
mnist & 784 & 0.99 & 1.67 & 2.09 \\
\hline
\end{tabular}
\end{center}
\caption{Empirically calculated tree imbalances, normalized by $N$.}
\label{tab:imbalance}
\end{table}

>>>>

\begin{figure}[b!]
\begin{subfigure}[b]{0.585\textwidth}
  \begin{center}
    \input{figures/balanced-cover-tree.tex}
  \end{center}
  \caption{Balanced cover tree.}
  \label{fig:imbalance-good}
\end{subfigure}
\begin{subfigure}[b]{0.415\textwidth}
  \begin{center}
    \input{figures/imbalanced-cover-tree.tex}
  \end{center}
  \caption{Imbalanced cover tree.}
  \label{fig:imbalance-bad}
\end{subfigure}
\caption{Balanced and imbalanced cover trees.}
\label{fig:imbalance}
\end{figure}

@@

\begin{defn}
The {\it cover node imbalance} $I_n(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf (that is, $s_i = -\infty$), then the number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $I_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $I_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$.\footnote{Note that in
this situation, $c \sim N$ also.}

However, for most real-world datasets with the cover tree implementation in {\bf
mlpack} \citep{mlpack2013} and the reference implementation
\citep{langford2006}, the tree imbalance is near-linear with the number of
points.  We have constructed cover trees on $N$ uniformly subsampled points
from a variety of datasets and calculated the imbalance; see Table
\ref{tab:imbalance} for the results.  Ten trials were performed for each dataset
and each $N$, and the mean imbalance is given.  These results are normalized
with respect to $N$, for which the values of $5000$, $50000$, and $500000$ were
chosen.  The `power', `susy', `higgs', and `covertype' datasets are found in the
UCI Machine Learning Repository \citep{ucimlrepository}, the `mnist' dataset is
from \cite{lecun2000mnist}, the `lcdm` and `sdss' datasets are Sloan Digital Sky
Survey data \citep{adelman2008sixth}, and the `randu' dataset is
randomly-generated uniformly-distributed data in 10 dimensions.  The imbalances
on each of these datasets tend to be near-linear.

@@

\begin{table}[tb]
\begin{center}
\begin{tabular}{| c | c | c | c | c |}
\hline
 & & \multicolumn{3}{|c|}{Imbalance} \\
\hline
Dataset & $d$ & $N = 5k$ & $N = 50k$ & $N = 500k$ \\
\hline
%lcdm & 3 & 22417.9 & 257445.6 & 2621239.6 \\
lcdm & 3 & 4.48 & 5.15 & 5.24 \\
% sdss & 4 & 10853.6 & 140431.6 & 1486808.8 \\
sdss & 4 & 2.17 & 2.81 & 2.97 \\
% power & 7 & 27042.2 & 323125.6 & 2251317.2
power & 7 & 5.41 & 6.46 & 4.50 \\
% susy & 18 & 3701.3 & 37823.7 & 431784.8 \\
susy & 18 & 0.74 & 0.76 & 0.86 \\
%randu-10 & 10 & 1156.0 & 10763.4 & 297041.2 \\
randu & 10 & 0.23 & 0.22 & 0.59 \\
%higgs & 29 & 4963.6 & 84171.8 & 781268.7 \\
higgs & 29 & 0.99 & 1.68 & 1.56 \\
%covertype & 54 & 6609.6 & 88296.4 & 1247820.7 \\
covertype & 54 & 1.322 & 1.766 & 2.495 \\
% mnist & 784 & 4974.1 & 83658.6 & 1044075.0 \\
mnist & 784 & 0.99 & 1.67 & 2.09 \\
\hline
\end{tabular}
\end{center}
\caption{Empirically calculated tree imbalances, normalized by $N$.}
\label{tab:imbalance}
\end{table}

### 2015-10-30 bounds-files/sec02-nearest-neighbor-search.tex
Particularly trivial revisions.

Given a type of tree and traversal, these two functions store the current
nearest neighbor candidates in the array $N$ and their distances in the array
$D$. \citep[See][for a more complete
discussion of how this algorithm works and a proof of
correctness.]{curtin2013tree}  The
\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
represents the maximum distance that could possibly improve a nearest neighbor
candidate for any descendant point of the query node $\mathscr{N}_q$.  The
standard bound function $B(\mathscr{N}_q)$ used for cover trees is adapted from
\citep{langford2006}:

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
I_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $I_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current tightest worst-case runtime bound for nearest
neighbor search.

>>>>

Given a type of tree and traversal, these two functions store the current
nearest neighbor candidates in the array $N$ and their distances in the array
$D$. \citep[See][for a more complete
discussion of how this algorithm works and a proof of
correctness.]{curtin2013tree}  The
\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
represents the maximum distance that could possibly improve a nearest neighbor
candidate for any descendant point of the query node $\mathscr{N}_q$.  The
standard bound function $B(\mathscr{N}_q)$ used for cover trees is adapted from
\citet{langford2006}:

@@

In the monochromatic case where $S_q = S_r$\footnote{In the monochromatic case,
we do not take a point as its own nearest neighbor, so slight modification of
{\tt BaseCase()} is necessary.  The runtime bound result remains unchanged.},
the bound is $O(c^9 (N +
I_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $I_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current tightest worst-case runtime bound for nearest
neighbor search.
